---
title: "TIME SERIES DATA ANALYSIS"
author: "RimaZinjuwadia"
date: "`r Sys.Date()`"
output: html_document
---



```{r message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Required libraries

library(tidyverse)    
library(lme4)         
library(ez)           
library(performance)  
library(cluster)      
library(ggplot2)      
library(dplyr)        
library(lmerTest)     
library(purrr)
library(tibble)
library(ggVennDiagram)
library(tidyr)
```

## Results {.tabset}


### 1) SUMMARY OF DATA & MODELING FRAMEWORK

The dataset analyzed in this study comprises protein abundance measurements from Saccharomyces cerevisiae subjected to osmotic stress, as originally published by Selevsek et al. (2015). It includes 18 samples collected across six time points (T00, T015, T030, T060, T090, and T120), with three biological replicates per time point. 

The analysis was based on two primary input files:

1) Selevsek2015_DIA_Spectronaut_annotation.csv, which contained metadata including experimental conditions and biological replicate identifiers
2) Selevsek2015.csv, which included quantified protein abundance values for each sample run  

The study followed a single factor repeated measures design and employed univariate statistical methods to evaluate both individual protein responses and overall expression patterns. The models applied included Single-Factor ANOVA Model, Single-Factor Repeated measure ANOVA, a Linear Mixed-Effects (LMM), and Linear Model (fallback).


```{r}
# Metadata contains sample annotations -> condition(time points - 6) , (replicate - 3)
meta <- read.csv("Selevsek2015_DIA_Spectronaut_annotation.csv") 
proteins <- read.csv("selvesek2015.csv") # contains the protein abundance 

# Merging meta by matching originalRUN
df <- proteins %>%
  pivot_longer(-originalRUN, names_to = "Protein", values_to = "Abundance") %>%
  left_join(meta, by = c("originalRUN" = "Run")) %>%
  filter(!is.na(Abundance))
print(df)
```


### 2) DATA PREPROCESSING & TRANSFORMATION

###### a)  

To prepare the data for statistical analysis, the protein abundance matrix was reshaped from wide to long format, facilitating the application of linear and repeated measures models. The reshaped data were then merged with the annotation metadata using run identifiers to generate a unified, analysis-ready dataset. Proteins with missing values or low variability (defined as standard deviation ≤ 0.01) were removed to ensure statistical robustness.


```{r}
# Load Metadata And Protein Abundance Data
meta <- read.csv("Selevsek2015_DIA_Spectronaut_annotation.csv")    
proteins <- read.csv("selvesek2015.csv")                            

# Reshape And Merge: Convert Wide To Long Format, Then Add Metadata
df <- proteins %>%
  pivot_longer(-originalRUN, names_to = "Protein", values_to = "Abundance") %>%  # Tidy long format
  left_join(meta, by = c("originalRUN" = "Run")) %>%                             # Join metadata
  filter(!is.na(Abundance))                                                      # Remove missing measurements

# Print and Save Data File
print(df)
write.csv(df, "merged_protein_long_format.csv", row.names = FALSE)
```
###### b)  

From the filtered dataset, a random subset of 150 proteins was selected to support both exploratory analysis and efficient univariate modeling.

```{r}
# Randomly Select 150 Proteins
set.seed(123)
selected_proteins <- sample(unique(df$Protein), 150)
print(selected_proteins)

# Subset Data By Selected Proteins
df_subset <- df %>% filter(Protein %in% selected_proteins)
print(df_subset)

# Save the subset data
write.csv(df_subset, "subset_protein_data.csv", row.names = FALSE)
```


### 1) SINGLE-FACTOR ANOVA MODEL

```{r}
# Create Dataframe to store anova p-values
aov_results <- data.frame()
for (prot in unique(df_subset$Protein)) {
  temp <- df_subset %>% filter(Protein == prot)
  pval <- tryCatch({
    summary(aov(Abundance ~ Condition, data = temp))[[1]]$`Pr(>F)`[1]
  }, error = function(e) NA)
  aov_results <- rbind(aov_results, data.frame(Protein = prot, AOV_p = pval))
}
write.csv(aov_results, "results_single_factor_anova.csv", row.names = FALSE)
print(aov_results)
```


##### a)  Distribution of One-way ANOVA p-values 

```{r}
# Distribution of p-values into standard significance thresholds.
aov_results$Significance <- cut(
  aov_results$AOV_p,
  breaks = c(-Inf, 0.001, 0.05, Inf),
  labels = c("< 0.001", "0.001 – 0.05", "> 0.05"),
  right = FALSE
)
# Create summary as tibble
summary_tbl <- as.data.frame(table(Significance = aov_results$Significance)) %>%
  as_tibble() %>%
  rename(Count = Freq) %>%
  mutate(Percentage = round(Count / sum(Count) * 100, 1))
# Print to console
print(summary_tbl)
write.csv(summary_tbl,"(summary_tbl.csv", row.names = FALSE)
```

##### b) Per Protein One-Way Anova Table

```{r}
aov_full_list <- list()
for (prot in unique(df_subset$Protein)) {
  temp <- df_subset %>% filter(Protein == prot)
  result <- tryCatch({
    model <- aov(Abundance ~ Condition, data = temp)
    tbl <- as.data.frame(summary(model)[[1]])
    tbl$Protein <- prot
    tbl$Term <- rownames(tbl)
    tbl
  }, error = function(e) NULL)

  if (!is.null(result)) {
    aov_full_list[[prot]] <- result
  }
}
aov_full_df <- do.call(rbind, aov_full_list)
# Round or format values
aov_full_df <- aov_full_df %>%
  mutate(across(
    c(`Sum Sq`, `Mean Sq`, `F value`, `Pr(>F)`),
    ~ round(.x, 3)
  ))
aov_full_df <- aov_full_df[, c("Protein", "Term", "Df", "Sum Sq", "Mean Sq", "F value", "Pr(>F)")]
# Save to CSV
write.csv(aov_full_df, "anova_full_table_per_protein.csv", row.names = FALSE)
# Preview
head(aov_full_df)
```



### REPEATED MEASURES ANOVA MODEL


##### a) RM ANOVA P-values

```{r}
rm_results <- data.frame()
for (prot in unique(df_subset$Protein)) {
  temp <- df_subset %>%
    filter(Protein == prot) %>%
    mutate(
      Condition = as.factor(Condition),
      BioReplicate = as.factor(BioReplicate)
    )
  if (length(unique(temp$Condition)) < 2 || length(unique(temp$BioReplicate)) < 2) {
    rm_results <- rbind(rm_results, data.frame(Protein = prot, RM_p = NA))
    next
  }
  # Repeated Measures ANOVA
  pval <- tryCatch({
    # within-subject correlation
    model <- aov(Abundance ~ Condition + Error(BioReplicate/Condition), data = temp) 
    # Extract p-value manually
    summary_out <- summary(model)
    p <- summary_out[[2]][[1]]$`Pr(>F)`[1]
    p
  }, error = function(e) {
    NA
  })
  rm_results <- rbind(rm_results, data.frame(Protein = prot, RM_p = pval))
}

#write.csv(rm_results, "results_repeated_measures_ANOVA.csv", row.names = FALSE)
print(rm_results)
```

##### b) RM ANOVA P-values & F-values Summary and BioreRlicate count check

```{r}
library(dplyr)
rm_results <- data.frame()
replicate_summary <- data.frame()  # track replicate counts
for (prot in unique(df_subset$Protein)) {
  temp <- df_subset %>%
    filter(Protein == prot) %>%
    mutate(
      Condition = as.factor(Condition),
      BioReplicate = as.factor(BioReplicate)
    )
  # Count replicates and timepoints
  num_reps <- length(unique(temp$BioReplicate))
  num_conds <- length(unique(temp$Condition))
  # Store replicate info
  replicate_summary <- rbind(replicate_summary, data.frame(
    Protein = prot,
    Num_Replicates = num_reps,
    Num_Timepoints = num_conds
  ))
  # Skip if data is incomplete
  if (num_conds < 2 || num_reps < 2) {
    rm_results <- rbind(rm_results, data.frame(
      Protein = prot,
      RM_p = NA,
      F_value = NA
    ))
    next
  }
  # Run RM ANOVA
  pval <- NA
  fval <- NA
  tryCatch({
    model <- aov(Abundance ~ Condition + Error(BioReplicate/Condition), data = temp)
    summary_out <- summary(model)
    pval <- summary_out[[2]][[1]]$`Pr(>F)`[1]
    fval <- summary_out[[2]][[1]]$`F value`[1]
  }, error = function(e) {})
  rm_results <- rbind(rm_results, data.frame(
    Protein = prot,
    RM_p = pval,
    F_value = fval
  ))
}
# Save results
#write.csv(rm_results, "results_rm_anova_only.csv", row.names = FALSE)
#write.csv(replicate_summary, "replicate_summary_check.csv", row.names = FALSE)

# View summaries
print(head(rm_results))
print(head(replicate_summary))


# Categorize p-values into bins
rm_results$P_Significance <- cut(
  rm_results$RM_p,
  breaks = c(-Inf, 0.001, 0.05, Inf),
  labels = c("< 0.001", "0.001 – 0.05", "> 0.05"),
  right = FALSE
)

# Categorize F-values into bins
rm_results$F_Range <- cut(
  rm_results$F_value,
  breaks = c(-Inf, 1, 2, 5, 10, Inf),
  labels = c("< 1", "1 – 2", "2 – 5", "5 – 10", "> 10"),
  right = FALSE
)

# Summary table for p-value bins
pval_summary_tbl <- as.data.frame(table(Significance = rm_results$P_Significance)) %>%
  as_tibble() %>%
  rename(Count = Freq) %>%
  mutate(Percentage = round(Count / sum(Count) * 100, 1))

# Summary table for F-value bins
fval_summary_tbl <- as.data.frame(table(F_Value_Bin = rm_results$F_Range)) %>%
  as_tibble() %>%
  rename(Count = Freq) %>%
  mutate(Percentage = round(Count / sum(Count) * 100, 1))

# Print results
print(pval_summary_tbl)
print(fval_summary_tbl)

# Save results
write.csv(fval_summary_tbl, "results_fval_summary_tbl.csv", row.names = FALSE)
write.csv(pval_summary_tbl, "results_pval_summary_tbl.csv", row.names = FALSE)

```

##### c) RM ANOVA Summary per protein

```{r}
# Initialize result list
rm_results_full <- data.frame()

# Loop through proteins
for (prot in unique(df_subset$Protein)) {
  temp <- df_subset %>%
    filter(Protein == prot) %>%
    mutate(
      Condition = as.factor(Condition),
      BioReplicate = as.factor(BioReplicate)
    )
  
  # Skip if insufficient groups
  if (length(unique(temp$Condition)) < 2 || length(unique(temp$BioReplicate)) < 2) {
    rm_results_full <- bind_rows(rm_results_full, 
                                 data.frame(Protein = prot, Term = NA, Df = NA, Sum_Sq = NA, Mean_Sq = NA, F_value = NA, P_value = NA))
    next
  }

  # Try RM ANOVA and extract the full table
  tryCatch({
    model <- aov(Abundance ~ Condition + Error(BioReplicate/Condition), data = temp)
    summary_out <- summary(model)

    # Extract BioReplicate stratum
    bio_stratum <- summary_out[["Error: BioReplicate"]][[1]]
    if (nrow(bio_stratum) > 0) {
      bio_replicate_row <- bio_stratum[1, ]
      rm_results_full <- bind_rows(rm_results_full,
        data.frame(
          Protein = prot,
          Term = "BioReplicate",
          Df = bio_replicate_row[["Df"]],
          Sum_Sq = bio_replicate_row[["Sum Sq"]],
          Mean_Sq = bio_replicate_row[["Mean Sq"]],
          F_value = bio_replicate_row[["F value"]],
          P_value = bio_replicate_row[["Pr(>F)"]]
        )
      )
    }

    # Extract Condition and Residuals from nested stratum
    cond_stratum <- summary_out[["Error: BioReplicate:Condition"]][[1]]
    rownames_stratum <- rownames(cond_stratum)

    for (i in seq_len(nrow(cond_stratum))) {
      row <- cond_stratum[i, ]
      rm_results_full <- bind_rows(rm_results_full,
        data.frame(
          Protein = prot,
          Term = rownames_stratum[i],
          Df = row[["Df"]],
          Sum_Sq = row[["Sum Sq"]],
          Mean_Sq = row[["Mean Sq"]],
          F_value = row[["F value"]],
          P_value = row[["Pr(>F)"]]
        )
      )
    }
  }, error = function(e) {
    rm_results_full <<- bind_rows(rm_results_full,
                                  data.frame(Protein = prot, Term = NA, Df = NA, Sum_Sq = NA, Mean_Sq = NA, F_value = NA, P_value = NA))
  })
}

# Round numeric values for better readability
rm_results_full <- rm_results_full %>%
  mutate(across(
    c(Sum_Sq, Mean_Sq, F_value, P_value),
    ~ round(.x, 3)
  ))

# Save to CSV
write.csv(rm_results_full, "rm_anova_full_results.csv", row.names = FALSE)

# View a few results
print(head(rm_results_full))
```

##### d) RM ANOVA Summary

```{r}
summary_stats_clean <- tibble::tibble(
  Metric = c(
    "Total Proteins Tested",
    "Valid p-values", 
    "Significant p-values (p < 0.05)", 
    "Highly Significant p-values (p < 0.001)",
    "Mean p-value", 
    "Median p-value", 
    "Min p-value", 
    "Max p-value",
    "Mean F-value", 
    "Median F-value", 
    "Max F-value"
  ),
  Value = c(
    nrow(rm_results),
    sum(!is.na(rm_results$RM_p)),
    sum(rm_results$RM_p < 0.05, na.rm = TRUE),
    sum(rm_results$RM_p < 0.001, na.rm = TRUE),
    round(mean(rm_results$RM_p, na.rm = TRUE), 4),
    round(median(rm_results$RM_p, na.rm = TRUE), 4),
    round(min(rm_results$RM_p, na.rm = TRUE), 4),
    round(max(rm_results$RM_p, na.rm = TRUE), 4),
    round(mean(rm_results$F_value, na.rm = TRUE), 4),
    round(median(rm_results$F_value, na.rm = TRUE), 4),
    round(max(rm_results$F_value, na.rm = TRUE), 4)
  )
)

print(summary_stats_clean)

```

### 3 & 4) Linear Mixed Effects Model (LMM) & Linear Model (LM)

```{r message=FALSE, warning=FALSE}

#Fit LMM or fallback LM per protein

lmm_results <- function(df_subset) {
  tryCatch({
    if (length(unique(df_subset$Condition)) < 2 || length(unique(df_subset$BioReplicate)) < 2) {
      return(list(p = NA, aic = NA, icc = NA, method = "Insufficient"))
    }

    lmm <- lmer(Abundance ~ Condition + (1 | BioReplicate), data = df_subset, REML = FALSE)

    if (lme4::isSingular(lmm, tol = 1e-4)) {
      lm_fit <- lm(Abundance ~ Condition, data = df_subset)
      f_stat <- summary(lm_fit)$fstatistic
      f_pval <- pf(f_stat[1], f_stat[2], f_stat[3], lower.tail = FALSE)
      coefs <- summary(lm_fit)$coefficients
      cond_rows <- grep("^Condition", rownames(coefs))
      cond_pvals <- coefs[cond_rows, "Pr(>|t|)"]
      lm_p <- if (length(cond_pvals) > 0) min(cond_pvals, na.rm = TRUE) else NA
      return(list(p = lm_p, aic = AIC(lm_fit), icc = NA, method = "LM"))
    }

    coefs <- summary(lmm)$coefficients
    cond_rows <- grep("^Condition", rownames(coefs))
    cond_pvals <- coefs[cond_rows, "Pr(>|t|)"]
    lmm_p <- if (length(cond_pvals) > 0) min(cond_pvals, na.rm = TRUE) else NA
    lmm_aic <- AIC(lmm)

    vc <- VarCorr(lmm)
    var_subject <- as.numeric(vc$BioReplicate)
    var_resid <- attr(vc, "sc")^2
    icc <- var_subject / (var_subject + var_resid)

    list(p = lmm_p, aic = lmm_aic, icc = icc, method = "LMM")
  }, error = function(e) {
    list(p = NA, aic = NA, icc = NA, method = "Error")
  })
}

results <- map_dfr(unique(df_subset$Protein), function(prot) {
  df_prot <- df_subset %>% filter(Protein == prot)
  stats <- lmm_results(df_prot)
  tibble(
    Protein = prot,
    P_value = round(stats$p, 3),
    AIC = round(stats$aic, 2),
    ICC = round(stats$icc, 3),
    Method = stats$method
  )
})

write.csv(results, "final_model_selection_results.csv", row.names = FALSE)
print(head(results))

# Post-processing: Add p-values per method and model selection

# 1. Load or compute unified model results
final_results <- read.csv("final_model_selection_results.csv")

final_results <- final_results %>%
  mutate(
    LMM_p = ifelse(Method == "LMM", P_value, NA),
    LM_p  = ifelse(Method == "LM",  P_value, NA),
    LMM_AIC = ifelse(Method == "LMM", AIC, NA),
    LM_AIC  = ifelse(Method == "LM",  AIC, NA),
    Final_p = ifelse(!is.na(ICC) & ICC >= 0.01, LMM_p, LM_p),
    ModelUsed = ifelse(!is.na(ICC) & ICC >= 0.01, "LMM", "LM")
  )

# format numbers
fresults <- final_results %>%
  mutate(
    P_value = signif(P_value, 3),
    AIC = round(AIC, 2),
    ICC = round(ICC, 3),
    LMM_p = signif(LMM_p, 3),
    LM_p = signif(LM_p, 3),
    LMM_AIC = round(LMM_AIC, 2),
    LM_AIC = round(LM_AIC, 2),
    Final_p = signif(Final_p, 3)
  )


write.csv(fresults, "final_model_results.csv", row.names = FALSE)
print(head(fresults))

#Separate results by model type
results_LMM <- results %>% filter(Method == "LMM")
results_LM  <- results %>% filter(Method == "LM")

# Save each to a separate file
#write.csv(results_LMM, "results_LMM_only.csv", row.names = FALSE)
#write.csv(results_LM,  "results_LM_only.csv",  row.names = FALSE)
```


##### a) Venn Diagram of Statistically Significant Proteins Identified by LMM, LM, and Final Model Selection

```{r}
# Define significance flags and Venn diagram sets

final_results <- final_results %>%
  mutate(
    LMM_sig   = ModelUsed == "LMM" & !is.na(LMM_p) & LMM_p < 0.05,
    LM_sig    = ModelUsed == "LM"  & !is.na(LM_p)  & LM_p < 0.05,
    Final_sig = Final_p < 0.05
  )

sig_LMM   <- final_results %>% filter(LMM_sig) %>% pull(Protein)
sig_LM    <- final_results %>% filter(LM_sig)  %>% pull(Protein)
sig_Final <- final_results %>% filter(Final_sig) %>% pull(Protein)

venn_sets <- list(
  LMM   = sig_LMM,
  LM    = sig_LM,
  Final = sig_Final
)

venn_plot <- ggVennDiagram(venn_sets, label_alpha = 0, edge_size = 0.5) +
  scale_fill_gradient(low = "Gray", high = "White") +
  labs(
    title = "Venn Diagram of Significant Proteins by Model",
    subtitle = "Overlap between LMM, LM, and Final Selected Proteins (p < 0.05)"
  ) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 16, color = "black"),
    plot.subtitle = element_text(face = "bold", size = 12, color = "black"),
    text = element_text(color = "Black")
  )

print(venn_plot)


# Summary of significant protein overlaps (Venn diagram logic)
cat("Significant Proteins by Model:\n")
cat("  sig_LMM      :", length(sig_LMM), "\n")
cat("  sig_LM       :", length(sig_LM), "\n")
cat("  sig_Final    :", length(sig_Final), "\n")
cat("  LMM ∩ Final  :", length(intersect(sig_LMM, sig_Final)), "\n")
cat("  LM ∩ Final   :", length(intersect(sig_LM, sig_Final)), "\n")
cat("  LMM ∩ LM     :", length(intersect(sig_LMM, sig_LM)), "\n")
cat("  LMM ∩ LM ∩ Final :", length(Reduce(intersect, list(sig_LMM, sig_LM, sig_Final))), "\n")

```

##### b) Summary Statistics for Proteins Modeled Using LMM & LMs

```{r}

# Summary: LMM and LM

lmm_summary <- final_results %>%
  filter(ModelUsed == "LMM", !is.na(LMM_p)) %>%
  summarize(
    Count_LMM       = n(),
    Significant_LMM = sum(LMM_p < 0.05, na.rm = TRUE),
    Mean_ICC        = mean(ICC, na.rm = TRUE),
    Max_ICC         = max(ICC, na.rm = TRUE),
    Mean_LMM_p      = mean(LMM_p, na.rm = TRUE),
    Median_LMM_p    = median(LMM_p, na.rm = TRUE)
  ) %>%
  mutate(
    Mean_ICC     = round(Mean_ICC, 3),
    Max_ICC      = round(Max_ICC, 3),
    Mean_LMM_p   = round(Mean_LMM_p, 3),
    Median_LMM_p = round(Median_LMM_p, 3)
  )

print(lmm_summary)


lm_summary <- final_results %>%
  filter(ModelUsed == "LM", !is.na(LM_p)) %>%
  summarize(
    Count_LM       = n(),
    Significant_LM = sum(LM_p < 0.05, na.rm = TRUE),
    Mean_LM_p      = mean(LM_p, na.rm = TRUE),
    Median_LM_p    = median(LM_p, na.rm = TRUE)
  ) %>%
  mutate(
    Mean_LM_p   = round(Mean_LM_p, 3),
    Median_LM_p = round(Median_LM_p, 3)
  )

print(lm_summary)

# Optional: Save rounded LM summary
write.csv(lm_summary, "results_lm_summary.csv", row.names = FALSE)
write.csv(lmm_summary, "results_lmm_summary.csv", row.names = FALSE)
```


### 5) Reduced vs. Full Model Evaluation - Nested LM 

```{r}
# Nested LM Comparison(unbalanced design check)

anova_compare_results <- data.frame()

for (prot in unique(df_subset$Protein)) {
  temp <- df_subset %>% filter(Protein == prot)

  pval <- tryCatch({
    full_model <- lm(Abundance ~ Condition, data = temp) #Full
    null_model <- lm(Abundance ~ 1, data = temp) #Reduced 
    comp <- anova(null_model, full_model)
    comp$`Pr(>F)`[2]
  }, error = function(e) NA)

  anova_compare_results <- rbind(anova_compare_results, data.frame(Protein = prot, Nested_F_p = pval))
}

anova_compare_results <- anova_compare_results %>%
  mutate(
    Nested_F_p = round(Nested_F_p, 5),
    Significance = case_when(
      Nested_F_p < 0.001 ~ "< 0.001",
      Nested_F_p < 0.05  ~ "0.001 – 0.05",
      TRUE               ~ "> 0.05"
    ),
    Interpretation = ifelse(Nested_F_p < 0.05,
                            "Reject H₀: Condition has significant effect",
                            "Fail to reject H₀: No significant condition effect")
  )

# Model Hypothesis Summary
hypothesis_summary <- anova_compare_results %>%
  mutate(Decision = ifelse(Nested_F_p < 0.05, "Reject H₀", "Fail to reject H₀")) %>%
  group_by(Decision) %>%
  summarise(
    Count = n(),
    Percentage = round(n() / nrow(anova_compare_results) * 100, 1)
  )


#Print and save results

print(hypothesis_summary)
print(anova_compare_results)

write.csv(hypothesis_summary, "hypothesis_summary.csv", row.names = FALSE)
write.csv(anova_compare_results, "anova_compare_results.csv", row.names = FALSE)
```


### 6) MODEL USAGE ACROSS PROTEINS — All Models (ANOVA, RM ANOVA, LMM, LM, Nested LM)

```{r}
# Merge all models: ANOVA, RM ANOVA, LMM/LM (final_results), and Nested LM
merged_models <- aov_results %>%
  full_join(rm_results, by = "Protein") %>%
  full_join(final_results, by = "Protein") %>%
  full_join(anova_compare_results, by = "Protein")

# Add logical flags for model usage
model_flags <- merged_models %>%
  mutate(
    ANOVA      = !is.na(AOV_p),
    RM_ANOVA   = !is.na(RM_p),
    LMM        = ModelUsed == "LMM" & !is.na(Final_p),
    LM         = ModelUsed == "LM"  & !is.na(Final_p),
    NESTED_LM  = !is.na(Nested_F_p)
  ) %>%
  select(Protein, ANOVA, RM_ANOVA, LMM, LM, NESTED_LM)

# Reshape to long format for plotting
model_usage_long <- model_flags %>%
  pivot_longer(cols = -Protein, names_to = "Model", values_to = "Used") %>%
  filter(Used == TRUE)

# Plot heatmap
ggplot(model_usage_long, aes(x = Model, y = Protein)) +
  geom_tile(fill = "Gray") +
  theme_minimal(base_size = 14) +
  labs(
    title = "Protein-wise Model Usage (ANOVA, RM ANOVA, LMM, LM, Nested LM)",
    x = "Model",
    y = "Protein"
  ) +
  theme(
    text = element_text(color = "black"),        # Dark font for all text
    axis.text = element_text(color = "black"),   # Dark font for axis labels
    axis.title = element_text(color = "black"),  # Dark font for axis titles
    plot.title = element_text(color = "black"),  # Dark font for plot title
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid = element_blank()
  )


# Optional: Print model usage summary table
model_summary_table <- model_usage_long %>%
  group_by(Model) %>%
  summarise(Count = n(), .groups = "drop")

print(model_summary_table)

```

##### a) MODEL USAGE ACROSS PROTEINS — All Models (ANOVA, LMM, LM)

```{r}

# Merge ANOVA, LMM, and LM results only (exclude RM)
merged_models <- aov_results %>%
  full_join(final_results, by = "Protein") %>%   # final_results includes LMM and LM outcomes
  full_join(lm_results, by = "Protein")

# Add Logical Flags for Model Usage
model_flags <- merged_models %>%
  mutate(
    ANOVA = !is.na(AOV_p),
    LMM   = ModelUsed == "LMM" & !is.na(Final_p),
    LM    = ModelUsed == "LM"  & !is.na(Final_p)
  ) %>%
  select(Protein, ANOVA, LMM, LM)

# Reshape to Long Format for Plotting
model_usage_long <- model_flags %>%
  pivot_longer(cols = -Protein, names_to = "Model", values_to = "Used") %>%
  filter(Used == TRUE)

# Plot Heatmap of Model Usage
ggplot(model_usage_long, aes(x = Model, y = Protein)) +
  geom_tile(fill = "Gray") +
  theme_minimal(base_size = 14) +
  labs(
    title = "Protein-wise Model Usage (ANOVA, LMM, LM)",
    x = "Model",
    y = "Protein"
  ) +
  theme(
    axis.text.y = element_blank(),        # Hide protein names if too many
    axis.ticks.y = element_blank(),
    panel.grid = element_blank()
  )
```


### 7) Pairwise Comparisons

##### a) Tukey Pairwise All Proteins

```{r}
# Prepare A Results List To Collect All Comparisons
tukey_all <- list()

for (prot in unique(df_subset$Protein)) {
  temp <- df_subset %>% filter(Protein == prot)
  model <- tryCatch({
    aov(Abundance ~ Condition, data = temp)
  }, error = function(e) NULL)

  if (!is.null(model)) {
    tukey <- tryCatch({
      as.data.frame(TukeyHSD(model, "Condition")$Condition)
    }, error = function(e) NULL)

    if (!is.null(tukey)) {
      tukey$Protein <- prot
      tukey$Comparison <- rownames(tukey)
      tukey_all[[prot]] <- tukey
    }
  }
}

# Combine All Results Into One Dataframe
tukey_df <- bind_rows(tukey_all)

# Apply FDR Correction To P-values
tukey_df$adj_p <- p.adjust(tukey_df$`p adj`, method = "fdr")

# Save to CSV
write.csv(tukey_df, "tukey_pairwise_all_proteins.csv", row.names = FALSE)
print(head(tukey_df, 10))  # Show first few results

```


##### b) TUKEY'S POST HOC PAIRWISE COMPARISONS (ONE-WAY ANOVA)

```{r}
# Prepare A Results List To Collect All Comparisons
tukey_all <- list()
for (prot in unique(df_subset$Protein)) {
  temp <- df_subset %>% filter(Protein == prot)
  model <- tryCatch({
    aov(Abundance ~ Condition, data = temp)
  }, error = function(e) NULL)
  if (!is.null(model)) {
    tukey <- tryCatch({
      as.data.frame(TukeyHSD(model, "Condition")$Condition)
    }, error = function(e) NULL)
    if (!is.null(tukey)) {
      tukey$Protein <- prot
      tukey$Comparison <- rownames(tukey)
      tukey_all[[prot]] <- tukey
    }
  }
}

# Combine All Results Into One Dataframe
tukey_df <- bind_rows(tukey_all)

# Applies FDR (Benjamini-Hochberg) Correction To P-values
tukey_df$adj_p <- p.adjust(tukey_df$`p adj`, method = "fdr")

# Save to CSV
write.csv(tukey_df, "tukey_pairwise_all_proteins.csv", row.names = FALSE)
print(tukey_df)
```



##### c) ESTIMATED MARGINAL MEANS/LEAST SQUARES MEANS PAIRWISE COMPARISON (LMM/LM MODEL)

```{r}
# Load Required Libraries
library(emmeans)
library(purrr)

# Load Dataset
df_subset <- read.csv("subset_protein_data.csv")
final_results <- read.csv("final_model_results.csv")

# Convert Factors
df_subset$Condition <- as.factor(df_subset$Condition)
df_subset$BioReplicate <- as.factor(df_subset$BioReplicate)

# Pre-Filter Good Proteins (3 Replicates Per Condition, All 6 Time Points)
good_proteins <- df_subset %>%
  dplyr::group_by(Protein, Condition) %>%
  dplyr::summarise(count = dplyr::n(), .groups = "drop") %>%
  dplyr::filter(count == 3) %>%
  dplyr::group_by(Protein) %>%
  dplyr::summarise(n_cond = dplyr::n_distinct(Condition), .groups = "drop") %>%
  dplyr::filter(n_cond == 6) %>%
  dplyr::pull(Protein)

cat("Number of proteins passing quality filter:", length(good_proteins), "\n")

# Run emmeans pairwise comparisons
emmeans_all <- list()

for (prot in good_proteins) {
  temp <- df_subset %>% filter(Protein == prot)
  model_type <- final_results$ModelUsed[final_results$Protein == prot]
  model <- tryCatch({
    if (model_type == "LMM") {
      lmer(Abundance ~ Condition + (1 | BioReplicate), data = temp)
    } else {
      lm(Abundance ~ Condition, data = temp)
    }
  }, error = function(e) {
    cat("Model failed for", prot, "\n")
    NULL
  })

  # Run emmeans if the Model Is Valid
  if (!is.null(model)) {
    result <- tryCatch({
      em <- emmeans(model, pairwise ~ Condition)
      out <- as.data.frame(em$contrasts)
      out$Protein <- prot
      out
    }, error = function(e) {
      cat("emmeans failed for", prot, "\n")
      NULL
    })
    if (!is.null(result)) {
      emmeans_all[[prot]] <- result
    }
  }
}

# Combine And Correct P-values
if (length(emmeans_all) == 0) {
  warning("No valid emmeans results were collected.")
  emmeans_df <- data.frame()
} else {
  emmeans_df <- bind_rows(emmeans_all)

  # Applies FDR (Benjamini-Hochberg) correction to p-values
  emmeans_df$adj_p <- p.adjust(emmeans_df$p.value, method = "fdr")

  # Save to CSV
  write.csv(emmeans_df, "emmeans_pairwise_all_proteins_filtered.csv", row.names = FALSE)
  print(head(emmeans_df))
}


```


##### d) PAIRWISE TIME POINT COMPARISONS VOLCANO PLOT 

```{r}
library(dplyr)
library(ggplot2)

# View available contrasts to confirm direction
print(unique(emmeans_df$contrast))  # Check for "T030 - T000" or "T000 - T030"

# Detect which contrast exists (adjust accordingly)
contrast_forward  <- "T030 - T000"
contrast_reverse  <- "T000 - T030"

# Filter for the available version of the contrast
if (contrast_forward %in% unique(emmeans_df$contrast)) {
  volcano_df <- emmeans_df %>%
    filter(contrast == contrast_forward & !is.na(adj_p)) %>%
    mutate(
      log2FC = estimate,  # T030 - T000
      neglog10_p = -log10(adj_p),
      Significance = ifelse(adj_p < 0.05, "FDR < 0.05", "NS")
    )
} else if (contrast_reverse %in% unique(emmeans_df$contrast)) {
  volcano_df <- emmeans_df %>%
    filter(contrast == contrast_reverse & !is.na(adj_p)) %>%
    mutate(
      log2FC = -estimate,  # Flip sign to get T030 - T000
      neglog10_p = -log10(adj_p),
      Significance = ifelse(adj_p < 0.05, "FDR < 0.05", "NS")
    )
} else {
  stop("Desired contrast (T030 vs T000) not found in emmeans_df$contrast.")
}

# Optional: inspect data
print(head(volcano_df))

# Create the volcano plot
ggplot(volcano_df, aes(x = log2FC, y = neglog10_p, color = Significance)) +
  geom_point(size = 2, alpha = 0.8) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
  scale_color_manual(values = c("FDR < 0.05" = "red", "NS" = "grey")) +
  labs(
    title = "Volcano Plot: Protein Abundance T030 vs T000",
    subtitle = "Based on model-based EMMeans pairwise comparison",
    x = "Log2 Fold Change (T030 – T000)",
    y = "-log10(FDR-adjusted p-value)"
  ) +
  theme_minimal() +
  theme(
    text = element_text(color = "black"),              # Make all text darker
    axis.text = element_text(color = "black"),         # Axis tick labels
    axis.title = element_text(color = "black"),        # Axis titles
    plot.title = element_text(color = "black", face = "bold"),    # Bold title
    plot.subtitle = element_text(color = "black")      # Subtitle text
  )
```



### 8) EXTRACT STATISTICALLY SIGNIFICANT PROTEIN DIFFERENCES

```{r}
# Load both files
tukey_df <- read.csv("tukey_pairwise_all_proteins.csv")
emmeans_df <- read.csv("emmeans_pairwise_all_proteins_filtered.csv")

# Filter for significant comparisons (adj_p < 0.05)
tukey_sig <- tukey_df %>% filter(adj_p < 0.05)
emmeans_sig <- emmeans_df %>% filter(adj_p < 0.05)

# Count Rejected Null Hypotheses 
cat("Tukey: Number of significant pairwise comparisons:", nrow(tukey_sig), "\n")
cat("Emmeans: Number of significant pairwise comparisons:", nrow(emmeans_sig), "\n")

# Preview Top Results
head(tukey_sig)
head(emmeans_sig)

# Save Significant Results
write.csv(tukey_sig, "tukey_significant_pairs.csv", row.names = FALSE)
write.csv(emmeans_sig, "emmeans_significant_pairs.csv", row.names = FALSE)

```

##### a) TOP 10 PROTEINS SIGNIFICANT COMPARISONS


```{r}
library(dplyr)
library(ggplot2)

# Step 1: Filter significant emmeans results
emmeans_sig <- emmeans_df %>% filter(adj_p < 0.05)

# Step 2: Create a label of significant contrasts per protein
emmeans_summary <- emmeans_sig %>%
  group_by(Protein) %>%
  summarise(
    n_sig_pairs = n(),
    timepoints = paste(contrast, collapse = ", "),
    .groups = "drop"
  )

# Step 3: Select Top 10 Proteins by number of significant contrasts
top_10 <- emmeans_summary %>%
  arrange(desc(n_sig_pairs)) %>%
  slice_head(n = 10)

# Step 4: Plot with timepoints as labels on the left side
ggplot(top_10, aes(x = reorder(Protein, -n_sig_pairs), y = n_sig_pairs)) +
  geom_col(fill = "Gray") +
  geom_text(
    aes(label = timepoints),
    hjust = 1.1, vjust = 0.5,
    angle = 0,
    color = "Black",
    size = 3.5,
    nudge_y = -0.1
  ) +
  labs(
    title = "Top 10 Proteins with Most Significant Pairwise Differences",
    x = "Protein",
    y = "Number of Significant Comparisons"
  ) +
  coord_flip() +  # Flip axes so labels are on the left
  theme_minimal(base_size = 14)

```






